---
title: "R and ggplot2"
output:
  html_document:
    df_print: paged
---

# R #

- Statistics-focused functional language
    - It looks more like JavaScript than Python
    - "Data scientists" will frequently swap between R and Python
- CRAN - open source repository for algorithms and libraries
    - The focal point for the statistics community in sharing code
    - Many advanced statistics algorithms are implemented in R
    - Good library support like *ggplot2* and *dplyr*
- RStudio, an IDE for *R*
    - Automatic integration with CRAN
    - Hyperlink help browser
    - Plot viewer and saved state integration
    - RMarkdown support, similar to Jupyter Notebooks
    - Shiny support, similar to Bokeh
- R + RStudio + CRAN is an example of a good software stack from an active, open-source community

---

# R Markdown

- Interactive documents or notebooks are done in Markdown.
    - Simplified "markup" language (hence the name Markdown)
    - For many sorts of documents: e.g., Github READMEs, web pages, etc.
- In RStudio, a cell (chunk) is part of the Markdown document, just like code in a Github README:

\`\`\`{r}

2 + 2

\`\`\`

`4`

- The chunk can be executed directly in the IDE
    - With *knitr*, directly create HTML, PDF, and Word
    - This section of the talk was done in RMarkdown

---

# Data Frames and R

- Fundamental support for data frames
- It also has support for databases and SQL, as well
    - DBI interface for databases: SQLite, PostgreSQL, etc.
    - Load tables using queries directly as data frames
- I won't spend much time showing data frames in R, as they are more or less the same as Python (Pandas)
- Rather, I'll show *dplyr*, a library/grammar for manipulating tables (relational data)

---

# CSV to Data Frame

We can directly load our cars CSV as a data frame:

```{r}
# read the data as a data frame
t <- read.csv("mtcars.csv") # assignment in R is <-
t
```

---

# Data Frame Selection

Just to show you that it's similar to Pandas, with slightly different syntax:

```{r}
# select rows 5-10, columns mpg and cyl
t[5:10, c('mpg','cyl')]
```

---

# Selecting Columns with *dplyr*

- Selecting data is done via operator chaining
    - aka functional combinators
    - aka data pipelines
    - aka data flow graphs
- `%>%` is the pipe operator
    - `|` in Unix
- *select* for picking columns:

```{r}
library(dplyr)

t %>% 
  select(name, mpg, cyl)
```

---

# Comparison to Bash Piping

In *bash*, it would look like:

```bash
cat mtcars.csv | awk -F',' '{ print $2, $3 }'
```

- Add in gnuplot
    - Another analysis and visualization environment
- See: <https://adamdrake.com/command-line-tools-can-be-235x-faster-than-your-hadoop-cluster.html>

---

# Filter (Where)

- In *SQL*
    - `WHERE` clause picks rows
- In *dplyr*
    - The `filter` operator

```{r}
# only show 3 gear cars, with their mpg and cyl
t %>% 
  filter(gear == 3) %>%
  select(name, mpg, cyl)
```

---

# Chaining More Filters

We can combine them further:

```{r}
# same query as before, but only keep greater than 200 hp
t %>%
  filter(gear == 3) %>%
  filter(hp > 200) %>%
  select(name, mpg, cyl)
```

---

# Adding in *Or*

We can use `or` in a filter, too:

```{r}
# same query as before, 3 or 4 gears
t %>%
  filter(gear == 3 || gear == 4) %>%
  filter(hp > 200) %>%
  select(name, mpg, cyl)
```

---

# Note for Databases and *dplyr*

- In these examples *t* is a data frame
    - In memory
- *dplyr* also works with databases
    - A reference to a table on disk, not in memory
    - Data are only copied into *R* on demand
    - Operate on Giga-/Tera-/Petabytes of data with *R*
- Lazy evalulation (or demand-driven)
    - Others: *Haskell*, *Spark*, *Dask*, *VTK*, etc.
    - Execution only happens when data are needed, like viewing or plotting

---

# Group By and Summaries

- `group_by` for `summarize`-ing (windowing) functions
  - min, max, average, etc.
  - *SQL* has these too
  
```{r}
# what's the average hp for each cylinder size?
t %>%
  group_by(cyl) %>%
  summarize(mean_hp = mean(hp))
```

---

# Sorting

- Sorting, `arrange`, is useful for line plots
    - *ORDER BY* in *SQL*

```{r}
# get cars with displacement < 100, and order by horsepower
t %>%
  filter(disp < 100) %>%
  select(name, hp, mpg) %>%
  arrange(hp)
```

---

# Plotting with ggplot2

- Created by Hadley Wickham
    - Any many other libraries, *dplyr* for example (*tidyverse*)
    - Inspired by "Grammar of Graphics" by Leland Wilkinson
- Works seamlessly with *dplyr*
    - Follows the data flow graph/piping paradigm
    - Pipe your data from *dplyr* directly into a plot
- Transformers/Operators manipulate the plot

---

# ggplot2 example

- *t* is filtered
- Then, "piped" directly into *ggplot*
    - *aes* and *geom_point* are operators on the plot
    - `+` is the "pipe" in *ggplot2*

```{r}
library(ggplot2)

# plot 3 gear cars, mpg vs. hp
t %>%
  filter(gear == 3) %>%
  select(mpg, hp) %>%
  ggplot(.) + # start a plot, . is placeholder for the incoming data
  aes(x=mpg, y=hp) + # aes, is "aesthetics" -- how to map data to geometry
  geom_point() # the plot geometry
```

---

# Shiny

- *Shiny* - interactive web graphics
    - Like *Bokeh* for Python
- Made of two parts
    - The UI - the interaction
    - The Server - the plot and data generator
- Unfortunately, doesn't work in static HTML, like *Bokeh*

```{r}
library(shiny)

# the front end, the UI
ui <- basicPage(
  plotOutput("plot1",
    hover = "plot_hover",
    brush = "plot_brush"
  ),
  verbatimTextOutput("info")
)

# the back end, the Server
server <- function(input, output) {
  # the plot
  output$plot1 <- renderPlot({
  # this is our plot that we had before
  t %>%
    filter(gear == 3) %>%
    select(mpg, hp) %>%
    ggplot(.) + 
    aes(x=mpg, y=hp) + 
    geom_point() 
  })

  # generating the tooltip text based on the user interaction
  output$info <- renderText({
    xy_str <- function(e) {
      if(is.null(e)) return("NULL\n")
      paste0("x=", round(e$x, 1), " y=", round(e$y, 1), "\n")
    }
    xy_range_str <- function(e) {
      if(is.null(e)) return("NULL\n")
      paste0("xmin=", round(e$xmin, 1), " xmax=", round(e$xmax, 1), 
             " ymin=", round(e$ymin, 1), " ymax=", round(e$ymax, 1))
    }

    paste0(
      "hover: ", xy_str(input$plot_hover),
      "brush: ", xy_range_str(input$plot_brush)
    )
  })
}

shinyApp(ui, server)
```
